---
title: "Introduction to Statistics for Parasitology"
format:
  html: 
    toc: true
    embed-resources: true
    number-sections: true
    fig-numbering: true
  pdf: 
    toc: false
    number-sections: true
    fig-numbering: true
editor: 
  markdown: 
    wrap: 72
---

# Overview

This document is an overview for some statistical methods commonly used
in parasitology (and biology in general). It is not meant to be a
statistical textbook, just a primer to get you up and running with your
own analysis. For this tutorial, you will need acomputer with a recent
version of [R and RStudio
installed](https://posit.co/products/open-source/rstudio). If you cannot
install RStudio, perhaps because you have a Chromebook, you can use the
Posit [RStudio Cloud
edition](https://posit.co/products/enterprise/cloud) (requires free
account signup).

::: callout-tip
Install R *before* installing RStudio. The download links at the link
above should take care of this for you.
:::

# Why do we have to do statistics?

Science works by testing **hypotheses**, or proposed explanations for
natural phenomena, against data. At some point, a scientist must
determine whether the data they have collected **support** or **do not
support** their hypothesis. Support or non-support manifest as the data
either matching the predictions of a hypothesis, or not matching the
predictions of a hypothesis. The **predictions** of a hypothesis are
patterns that would be expected in the data if the hypothesis were true.
In science, we require that any hypothesis be **falsifiable**, at least
in principle. This means that there must be some observation that would
prove a hypothesis false. That is, a false hypothesis must make at least
some false predictions. A true hypothesis, by contrast, should generate
*only* true predictions.

The trouble is, biological data are noisy. Individual organisms differ
from one another, measurements are imperfect, and random chance
influences every sample we collect. As a result, patterns can appear in
a dataset even when no real biological effect exists. Our goal is to
distinguish true biological **signals** from background **noise**.

Statistical tests help us do this. They ask whether the pattern we
observed is stronger than we would expect to see from random variation
alone. If the observed pattern would be very unlikely to arise from
noise by itself, we have evidence that a real signal may be present.
This evidence is summarized by the *p*-value, which estimates how
surprising our data would be if there were actually no biological
effect. Smaller *p*-values suggest a stronger signal relative to the
background noise, while larger *p*-values suggest that the observed
pattern could easily be explained by random variation. Patterns that are
stronger than what would be expected by chance are called
**statistically significant**. For historical (and somewhat arbitrary
and controversial reasons), biologists usually use $p<0.05$ as the
threshold for significance.

Let's consider an example. A researcher observes that she tends to find
more parasitic worms in larger fish. After some reading and thinking,
she develops a hypothesis:

**"Larger fish tend to have more parasites because larger fish tend to
be older, and thus have had more time to accumulate parasites."**

From this proposed explanation, the researcher can deduce several
predictions that logically follow:

1.  **Prediction 1:** Adult fish will have more parasites in their
    bodies than juvenile fish.

2.  **Prediction 2:** There will be a positive correlation between fish
    body length and the number of parasitic worms in that fish.

If both of these predictions are true, then that would be good support
for her hypothesis. If one or neither is true, then the hypothesis is
not supported and should either be revised or discarded.

The researcher collects data from as many fish as she can find, and gets
these results. First, the results related to prediction 1:

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4

set.seed(123)
d1 <- data.frame(age="Adult", y=rpois(100, 10))
d2 <- data.frame(age="Juvenile", y=rpois(100, 5))
dx <- rbind(d1, d2)

par(mfrow=c(1,1), mar=c(5.1, 5.1, 1.1, 1.1), bty="n", lend=1, las=1,
    cex.axis=1.4, cex.lab=1.4)
boxplot(y~age, data=dx, xlab="Host age", ylab="Parasite count", ylim=c(0, max(dx$y)))
```

Second, the results related to prediction 2:

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4


set.seed(321)
dy <- data.frame(x=runif(60, 5, 25))
dy$y <- rpois(60, 3+0.9*dy$x)
par(mfrow=c(1,1), mar=c(5.1, 5.1, 1.1, 1.1), bty="n", lend=1, las=1,
    cex.axis=1.4, cex.lab=1.4)
plot(dy$x, dy$y, xlab="Fish length (cm)", ylab="Parasite count",
     ylim=c(0, max(dy$y)))
```

Were her predictions supported? The figures suggest yes, but we need to
do a formal test in order to say yes with confidence.

For the first figure, what we want is to see if the means of the two
groups differ. If the means differ, then the hypothesis is supported. If
the means do not differ, then the hypothesis is *not* supported. It's
not enough to just look at the sample means--9.67 for adults and 4.92
for juveniles--we have to compare the pattern we observed to what could
have happened if the prediction was false. That is, if the means were
really the same. The proper test for this situation is called a
**Student's *t*-test**, or just ***t*-test**. In this case:

```{r}
t.test(y~age, data=dx)
```

The *p*-value indicates that the difference in means between the adults
and juveniles is **statistically significant:** larger and more
consistent than what would be expected if there was no difference
between groups.

In the case of prediction 2, the researcher could check for a **linear
correlation**, or reliable covariation between the "parasite count" and
"fish length":

```{r}
cor.test(dy$x, dy$y)
```

Again, the *p*-value less than 0.05 confirms the researcher's suspicion
that the correlation was stronger than might be expected if there was no
correlation.

These were just examples--we will explore each of these techniques in
detail below. For now, just remember:

1.  Scientists look for patterns, but data is noisy.

2.  Statistics helps us tell how likely it is we are seeing signal or
    noise--but does not **prove** whether a signal exists.

3.  A *p*-value can never be 0, so statistics can always lead us astray
    if we are not careful.

4.  "Statistically significant" is not the same as "biologically
    significant".

# Types of tests

There are thousands of statistical tests out there, but most biologists
only ever need a handful. One of the hardest parts of "doing statistics"
for many people is choosing which test to use. If you ask a
statistician, they are going to ask you, "Well, what *question* are you
trying to answer?". This is because every statistical test answers a
particular question:

-   Do groups A and B have the same mean?

-   As variable $x$ changes, does variable $y$ also tend to change?

-   If variable $x$ changes by this much, how much will $y$ change?

-   If event $G$ happens, how likely is it that event $H$ also happens?

-   Do values in group $A$ tend to be greater than, or smaller than,
    values in group $B$?

-   Are there categories that consistently and reliably capture the
    variation in these sample units?

Notice that all of the statements above are statistical statements, not
necessarily biological statements. Translating biological ideas and
questions into statistical statements and questions is one of the
hardest parts of doing statistics.

The hypotheses that we test in biology tend to fall into just a few
categories:

1.  Testing for a difference in **mean** or **location** between groups.

2.  Testing for a **continuous relationship** between a **response
    variable** (or multiple variables) and one or more **predictor
    variables**. Also referred to as predicting or **modeling** a
    continuous variable.

3.  Testing for (lack of) **independence between categories** of
    observations.

4.  **Classifying** observations based on predictor variables.

These categories are not mutually exclusive...some questions can be
analyzed in $>1$ way!

# Summary statistics

## Central tendency and variability

One of the first tasks in data analysis is to summarize the data. This
means presenting values that express the average or **central
tendency**--the typical value in a dataset--and some idea of the
**variability**.

Central tendency is usually expressed as the **mean** or **median**:

-   **Mean:** the sum of all values divided by the number of values.
    This tells you, "if all values were the same, what would they be?".

-   **Median:** the middle value or 50th percentile. If you lined up all
    values from smallest to greatest, this value would be in the center.

Variability is usually expressed as the **standard deviation** or
**interquartile range**:

-   **Standard deviation (SD):** the square-root of the **variance**, or
    mean squared deviation from the mean. The larger the SD, the more
    spread out the values tend to be.

-   **Interquartile range (IQR):** the difference between the 75th
    percentile and the 25th percentile of a set of observations. By
    definition, 50% of observations fall in this interval. The larger
    the IQR, the more spread out the value are.

When summarizing data, we usually present the sample mean $\bar{x}$ and
the sample SD $s$ as $\bar{x}\pm s$. E.g., $3.2 \pm 0.8$ cm.

We can see the effect of the mean and SD by examining some distributions with different $\mu$ and $\sigma$.

```{r}
# possible values of distribution
xvals <- seq(-5, 5, length=1000)

# calculate probability distribution function (PDF)
# of each x value under different mu and sigma. 
# PDF expresses relative likelihood of a value.
pdf1 <- dnorm(xvals, mean=0, sd=0.2)
pdf2 <- dnorm(xvals, mean=0, sd=1)
pdf3 <- dnorm(xvals, mean=0, sd=4)
pdf4 <- dnorm(xvals, mean=2, sd=0.5)

# plot the PDFs
plot(xvals, pdf1, type="l", lwd=3, 
     xlab="X value", ylab="PDF(x)",
     ylim=c(0, 2))
points(xvals, pdf2, type="l", lwd=3, col="red3")
points(xvals, pdf3, type="l", lwd=3, col="blue3")
points(xvals, pdf4, type="l", lwd=3, col="green4")
legend("topleft",
       legend=c(
         expression(mu==0~sigma==0.2),
         expression(mu==0~sigma==1),
         expression(mu==0~sigma==4),
         expression(mu==2~sigma==0.5)
       ),
       lwd=3, col=c("black", "red3", "blue3", "green4"))
```

The figure shows that the mean $\mu$ shifts the mean, which is also the most likely value (and thus peak of the graph). The SD $\sigma$ determines how spread out the values are.

## Sample vs. population statistics

When we collect data, we collect a **sample** of all possible values. The set of all possible values is the **population** of values. As biologists we are usually interested in making inferences about the population using estimates derived from sample data.

Let's explore some data from the [normal distribution](https://en.wikipedia.org/wiki/Normal_distribution) for an example. We will use R to draw *samples* from a distribution with a population mean ($\mu$) of 10 and a population SD ($\sigma$) of 3.

```{r}
# draw 10 values from the underlying distribution
rand.values <- rnorm(10, mean=10, sd=3)

# calculate the sample mean and sample SD
mean(rand.values)
sd(rand.values)
```

Your results will be different from those above, because the values are random, but the sample mean is likely around 10 and the sample SD around 3. If you repeat those 3 commands, you will get different numbers each time. The sample mean will probably never exactly equal the true $\mu$, but it will be close. If you take more samples, then the sample mean should become a better estimator of the population mean:

```{r}
# draw 100 values from the underlying distribution,
# then calculate the sample mean and SD
rand.values100 <- rnorm(100, mean=10, sd=3)
mean(rand.values100)
sd(rand.values100)
```

Because of randomness, your sample mean might be more correct with 10 samples than with 1000 samples! But, the long-run average of sample means will be better estimators of $\mu$ with larger sample sizes. We can test this with simulation:

```{r}
# population parameters
mu <- 10
sigma <- 3

# sample sizes to test
samp.sizes <- c(3:10, 2:10*10, 2:10*100)
n.samp.sizes <- length(samp.sizes)

# how many simulations per sample size?
sims <- 1000

# structure to store simulated sample means
# for each sample size
sim.list <- vector("list", length=n.samp.sizes)

# execute simulations in a loop
for(i in 1:n.samp.sizes){
  # vector to hold simulated sample means for
  # each replicate j
  sim.vector <- numeric(sims)
  # simulate sample means in a loop
  for(j in 1:sims){
    sim.vector[j] <- mean(rnorm(samp.sizes[i], mu, sigma))
  }#j
  # store in list
  sim.list[[i]] <- sim.vector
}#i

# what is the mean estimate at each sample size?
mean.mean <- sapply(sim.list, mean)
# what is the spread of estimates of mu at each
# sample size?
sim.ranges <- sapply(sim.list, range)

# plot the estimate and range of estimates vs.
# sample size
plot(samp.sizes, mean.mean, lwd=3, type="o", pch=16, log="x",
     cex=1.5, xlab="Sample size", ylab="Mean of sample means",
     ylim=c(5, 15))
segments(samp.sizes, sim.ranges[1,], samp.sizes, sim.ranges[2,],
         lwd=3)
```

As you can see, the estimates of the true population mean converge rather quickly on the correct value. But the uncertainty around that estimate (the width of the vertical bars) takes quite a while longer to shrink. Try repeating the code block above with different values of `mu` and `sigma` to see if the pattern changes.

## Summarizing data in R

The most common summary statistics have eponymous functions in R: `mean()`, `sd()`, `var()`, `IQR()`, `quantile()`, and so on. 

```{r}
#| collapse: true

# get some random values
# with mu = 0 and sigma = 1
# (standard normal distribution)
z <- rnorm(100)
```

```{r}
# mean
mean(z)
```

```{r}
# sd
sd(z)
```

```{r}
# variance
var(z)
```

```{r}
# quantiles 
## 20th percentile: 20% of values are <= this value
quantile(z, 0.2)
```

```{r}
## 90th percentile: 90% of values are <= this value
quantile(z, 0.9)
```

```{r}
# median
median(z)
```

```{r}
# also the median
quantile(z, 0.5)
```

```{r}
# IQR
IQR(z)
```

When summarizing values, we typically write either $\bar{x}\pm s$ or *median (IQR)*. Make sure not to confuse the sample mean $\bar{x}$ with the population mean $\mu$. Likewise, do not present the sample SD $s$ as the population SD $\sigma$.

:::{.callout-tip}
**Missing values**

When a dataset contains missing values, `NA`, you need to tell the summarizing function to ignore them. Usually we add the argument `na.rm=TRUE`. For example, `mean(x, na.rm=TRUE)`.
:::

## Summarizing by groups in R

To summarize data by group in R, use the function `aggregate()`. For this example, we will use the built-in *iris* dataset, [which contains measurements taken from 150 flowers of 3 different species of *Iris*](https://en.wikipedia.org/wiki/Iris_flower_data_set). 

We can use aggregate to make a table with the mean and SD of the petal length of each species in `iris`.

```{r}
# calculate mean of each species
ag.petal <- aggregate(Petal.Length~Species, data=iris, mean)

# calculate sd of each species
ag.petal$SD <- aggregate(Petal.Length~Species, data=iris, sd)$Petal.Length
ag.petal
```

Some people prefer to use `tapply()` instead of a second call to `aggregate()`:

```{r}
ag2 <- aggregate(Petal.Length~Species, data=iris, mean)
ag2$sd <- tapply(iris$Petal.Length, iris$Species, sd)
ag2
```

You can also use `tapply()` to build a data frame of summary statistics from scratch, although `aggregate()` is usually safer because of how it handles missing groups.

R function `aggregate()` does what Pivot Tables do in Excel: summarize by groups. However, R has a huge advantage over Excel in that it can summarize by any function you can define, as opposed to the short list of functions available in Excel. For example, you could count the number of values >= 3 or between arbitrary values like 3.5 and 4.7. A function defined and used within a function as shown below is called an **anonymous function**. 

```{r}
# count number of values >=3 with an anonymous function
ag3 <- aggregate(Petal.Length~Species, data=iris, 
                 function(x){length(which(x>=4))})
ag3
```

```{r}
# count number of values between 3.5 and 4.7
ag4 <- aggregate(Petal.Length~Species, data=iris,
                 function(x){length(which(x>3.5 & x<4.7))})
ag4
```

Here is an example where we count the number of missing values and number of infinite values...two very useful functions for figuring out why your code doesn't work! These examples also show the use of anonymous functions in `tapply()` as well as in `aggregate()`.

```{r}
tapply(iris$Petal.Length, iris$Species,
       function(x){length(which(is.na(x)))})
```

```{r}
aggregate(Petal.Length~Species, data=iris, 
          function(z){any(is.nan(z))})
```

:::{.callout-tip}
**which()-craft

In expressions like the one above, where we selected observations that met a certain condition (e.g., `which(x>4)`), the function `which()` is not strictly necessary. The expression 
:::

# Tests comparing groups

In a test for difference in mean or location, we are interested in
whether observations differ between groups. Most of the time we are testing whether the central tendency differs between groups. A test of this kind evaluates a prediction like, "a typical value in group A is different from a typical value in group B".

Alternatively, we can test for a difference in **location**: the
relative position of random values from 2 or more groups. This is a
slightly different statement: "values in group A tend to be greater than
(or less than) values in group B". Put another way: "If we pick a random
value from group A, it will be reliably greater (or less than) than a
random value from group B". Tests for location are performed on the **rank order** of the values, not the values themselves.

When comparing groups, the data are usually best shown using a
**boxplot**, also known as a **box-and-whiskers** plot. The exact
implementation varies between software packages, but boxplots usually
show the median, interquartile range (approximately), extreme values,
and possible outliers (if any) Run the command `?boxplot.stats` to see a description of how boxplots are calculated in R. 

## Comparing 2 groups: the *t*-test

The ***t*-test** compares the means of two groups. There are many kinds of *t*-tests, but the case of comparing two means is the most common. Technically, this is called a **"two-sided, two-sample *t*-test** or sometimes a **Student's *t*-test**, although the latter term is not accurate because modern software uses a different method of accounting for unequal variances between groups (the "Welch-Satterthwaite correction"). Most of the time you can just say *t*-test and people will know what you mean.

***t*-test example**

Here is an example of running a *t*-test on two samples. If you try the code yourself with a different random number seed (`set.seed()`), or without a random number seed, your numbers will be slightly different because of the random number generation.

```{r}
# make some data
set.seed(123) # ensures same numbers each time RNG is done
x1 <- rnorm(n=20, mean=10, sd=4)
x2 <- rnorm(n=20, mean=15, sd=4)

# run the test
t.test(x1, x2)
```

You can also use a *t*-test on data within a data frame using the formula interface. Notice that this code is essentially the same as the code for `boxplot()` and `aggregate()`.

```{r}
# grab 2 of the species in iris
use.species <- c("versicolor", "virginica")
iris.2spp <- iris[which(iris$Species %in% use.species),]

# remove the now-missing level of "setosa"
# (this will matter for the plot in the next code block)
iris.2spp$Species <- droplevels(iris.2spp$Species)

# run the t-test
t.test(Petal.Length~Species, data=iris.2spp)
```

Show the difference using a boxplot:

```{r}
boxplot(Petal.Length~Species, data=iris.2spp,
        xlab="Species", ylab="Petal length (cm)")
```

### Non-parametric alternative

Sometimes, your data will not meet all of the assumptions of a *t*-test. That is beyond the scope of what we are doing here, but if you find yourself in that situation, you can try a **Wilcoxon rank sum test** instead of  *t*-test. They syntax is the same as for `t.test()`.

```{r}
wilcox.test(Petal.Length~Species, data=iris.2spp)
```
## Comparing 3 or more groups: ANOVA

### One-way ANOVA

When there are 3 or more groups, we do not use a *t*-test and instead use **analysis of variance**, or **ANOVA**. There are two ways to do an ANOVA in R, one using `aov()` and the other using `lm()`. We'll use `aov()` 

When there is only 1 explanatory **factor** (grouping variable), regardless of how many **levels** (groups) it has, the test is called a **one-way ANOVA**.


```{r}
# aov style
mod1 <- aov(Petal.Length~Species, data=iris)
summary(mod1)
```

The `summary()` command produces the **ANOVA table**, which summarizes how the variance is partitioned between different sources. In this case, the total sum of squared deviations of all observations is $437.1 + 27.2=464.3$. ANOVA partitions that into variation due to species (437.1) and residual or "left over" variation (27.2). We sometimes think of the 437.1 as "between group" variation, and the 27.2 as "within group" variation. If there is a significant effect of species, then roughly speaking there should be more between group variation than within group variation. Really it is the variation per degree of freedom (`Df`) that defines that ratio, or $F$ statistic. Because of the way the table rounds values, it obscures the fact that $F=MSB/MSE$, or the ratio of the mean between group sums of squares and the mean within group sums of squares. The values shown, 281.55 and 0.19, are rounded so much that their ratio doesn't equal the printed F value.

Anyway, the **omnibus test** above shows that at least 1 level, or group, of `Species` differs from one of the others. To see the tests for the pairwise differences, we have to run a **post-hoc test**. One of the most common post-hoc tests for ANOVA is the Tukey Honest Significant Difference (HSD) test:

```{r}
TukeyHSD(mod1)
```

The Tukey output shows us that every group differs from every other, with *p<0.001* (here *p* is rounded to 0, although it can never be 0). The differences in group means are presented with their **95% confidence intervals (CI)**. For example, the difference between the group means of versicolor and setosa is 2.798 cm, with 95% CI = [2.594, 3.002]. 

### Two-way ANOVA (and more way)

When there are 2 or more explanatory factors, the test is called a **two-way ANOVA**, or **three-way ANOVA**, and so on. We can add additional factors to an ANOVA with the `+` sign.

```{r}
# make a copy of iris with an extra grouping variable
iris2 <- iris
iris2$Flower <- c("purple", "pink")

# fit the two-way ANOVA without interaction
mod2 <- aov(Petal.Length~Species+Flower, data=iris2)
summary(mod2)
```

Adding the colors that way just repeated them until the column was full: purple, pink, purple, pink, and so on until the new vector had length 150 (the number of rows in `iris`). This essentially distributes the colors randomly with respect to the species and thus the petal lengths, so it should be no surprise that flower color had no effect on petal length ($p=0.61$). A Tukey test confirms the effects of species, and the non-effect of `Flower`. Technically, you don't need to do or report a Tukey test if the main effect is nonsignificant.

```{r}
TukeyHSD(mod2, which="Species")
```

Something we could also ask is whether there was an **interaction** between species and color: a situation where one variable changes the effect of another. Maybe purple an pink flowers really do have different petal lengths, but only in one species. Testing for an interaction would help detect that kind of pattern. To test for an interaction between two variables, use the `*` symbol instead of the `+`.

```{r}
# fit the two-way ANOVA WITH interaction
mod3 <- aov(Petal.Length~Species*Flower, data=iris2)
summary(mod3)
```

The interaction is nonsignificant, so we should instead report the main effects in the model without the interaction (`mod2`). If there is an interaction present, then it becomes impossible to discuss the effects of one variable without discussing the effects of the other. We will explore interactions more in the section on ANCOVA below.

## Nonparametric alternatives

# Correlation

## Linear correlation

## Nonlinear correlation

## Do you need a *p*-value for a correlation?

# Linear models

## Linear regression

## Analysis of covariance (ANCOVA)

# Generalized linear models

## Models for counts: Poisson GLM

## Models for presence: Logistic Regression

## Models for proportions: Binomial Regression
